Plant Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging

### 2022-12-04 script/manuscript_MEE.Rmd
more writing in intro, more metabarcoding content added

The inability to reliably identify plants to the level of species often leaves our understanding of ecosystem function and interactions wanting. Current methods to ameliorate this situation include: ignoring these ecologically relevant levels of detail, revisiting plots as diagnostic material becomes temporally available, assistance from taxonomic specialists, or the use of barcoding or other molecular techniques. These approaches are untenable in light of the benefits offered by: species in several morphologically difficult genera which serve as bioindicators, preferred partners in ecological interactions, as well as an increasing lack of taxonomic experts (CITE). Many genera, especially with the formalized advent of integrative taxonomy, have species which are well defined based upon ecological and behavioral rather than morphological properties, the identification of these taxa in degraded areas or without their mutualistic partners is fraught with difficulty. Hindering an understanding of the breadth of habitat which some species occupy, and the interactions they have with other species.

  The identification of many plant species to terminal taxon is an essential component of nearly all land management programs, where many species in the same genus  (e.g. Sagebrush - *Artemisia* L., Willows - *Salix* L., and Sedges - *Carex* L.) serve as bioindicators (respectively for 'rangelands', streams, and wetlands), as well as in academic research (@Gage2013HistoricalRO, AIM). This endeavour is often mired by lack of diagnostic characters (e.g. flowers, fruits, or roots),  and increasingly the description of cryptic species (@janzen2017nuclear, @oliver2009cryptic). Solutions to this problem are wanting, certain programmes have relied increasingly upon revisiting field sites to identify material using morphological or chemical approaches, whereas academic research has often used high copy number plastid genes as barcodes (Rosentreter et al. 2021, MORE MORE). However, both approaches have significant downsides, the former resource intensive at the landscape scale - and often does not work, while the latter seldom works due to a lack of variability in the currently available barcodes (@liu2014identification).

  Recently barcoding, and metabarcoding, have shown considerable promise in all Kingdoms of life. For example …. . With plants the identification of members of certain clades has been quite successful, whereas with others results have been elusive (@liu2014identification), most applications lay along this spectrum (CITE). Particular challenges with the utilization of high-copy number sequences are associated with their rates of divergence.... and... Herein we have resolved major components of the problems of identifying plant material without diagnostic morphological character states using the Angiosperms353 (A353) Hyb-Seq probes (@johnson2019universal), and custom species sequence databases derived via species distribution modelling, and temporal filtering.

@@

  Currently the largest plant systematic endeavor ever undertaken, the Kew Plant and Fungal Tree of Life (PAFTOL), is approaching completion (@baker2021PAFTOL). This dataset will contain Hyb-Seq data from at least one species representing each genus in the plant kingdom using the popular A353 probes (@baker2021exploring), resulting in over 14,000 represented species. These publically available data serve to provide a taxonomically comprehensive backbone for plant metabarcoding. Data from the 10kP project, which seeks to develop reference genomes from a phylogenetically diverse suite of plants will contribute many more records upon it's intended completion, now slated to be by 2030, similar projects which seek to sequence high amounts of genomes in regions e.g. the 'Darwin Tree of Life' are being undertaken which will contribute data for applicable to enormous spatial domains (@cheng2018tenkp, @darwin2022project, @lewin2022biogenome).

@@

## 3.2 | Microscopic Pollen identification

@@

Only 58 of these 388 species (n = 34.56897, Mdn n = 31) were able to be compared to ground truth data from long term (1970 ? - 2016 ?) datasets.

@@

bee_obs <- read.csv(
  file.path('../data', files[grep('queen_observations', files)]) ) %>% 
  filter(caste == 'q') %>% 
  mutate(species = gsub('\\..*$', '', species)) %>% 
  select(site, doy, week, tot.obs.length, species, plant.species,
         resource.coll:pollen.collected, fl.switch.1:fl.switch.2.resource.coll) %>% 
  mutate(across(resource.coll:fl.switch.2.resource.coll, ~ na_if(.x, ''))) %>% 
  pivot_longer(c(plant.species, fl.switch.1, fl.switch.2), names_to = 'observation', values_to = 'plant.species') %>% 
  drop_na(plant.species) %>% 
  group_by(site, week, species, plant.species) %>% 
  count(name = 'interactions') %>% 
  left_join(., obs_time) %>% 
  mutate(period = case_when(
    week %in% 3:5 ~ 'Early',
    week %in% 6:8 ~ 'Mid',
    week %in% 9:11 ~ 'Late'
  ))

@@

```{r Most used resources by species, eval = F}

>>>>

The inability to reliably identify plants to the level of species often leaves our understanding of ecosystem function and interactions wanting. Current methods to ameliorate this situation include: ignoring these ecologically relevant levels of detail, revisiting plots as diagnostic material becomes temporally available, assistance from taxonomic specialists, or the use of barcoding or other molecular techniques. These approaches are untenable in light of the benefits offered by: species in several morphologically difficult genera which serve as bioindicators, preferred partners in ecological interactions, as well as an increasing lack of taxonomic experts (@hebert2003biological). Many genera, especially with the formalized advent of integrative taxonomy, have species which are well defined based upon ecological and behavioral rather than morphological properties, the identification of these taxa in degraded areas or without their mutualistic partners is fraught with difficulty. Hindering an understanding of the breadth of habitat which some species occupy, and the interactions they have with other species.

  The identification of many plant species to terminal taxon is an essential component of nearly all land management programs, where many species in the same genus  (e.g. Sagebrush - *Artemisia* L., Willows - *Salix* L., and Sedges - *Carex* L.) serve as bioindicators (respectively for 'rangelands', streams, and wetlands), as well as in academic research (@Gage2013HistoricalRO, AIM). This endeavour is often mired by lack of diagnostic characters (e.g. flowers, fruits, roots or combinations thereof),  and increasingly the description of cryptic species (@janzen2017nuclear, @oliver2009cryptic). Solutions to this problem are wanting, certain programmes have relied increasingly upon revisiting field sites to identify material using morphological or chemical approaches, whereas academic research has often used high copy number plastid genes as barcodes (Rosentreter et al. 2021, MORE MORE). However, both approaches have significant downsides, the former resource intensive at the landscape scale - and often does not work, while the latter seldom works due to a lack of variability in the currently available barcodes (@liu2014identification).

  Recently barcoding, and metabarcoding, have shown considerable promise in all Kingdoms of life. For example …. . With plants the identification of members of certain clades has been quite successful, whereas with others results have been elusive (@liu2014identification, @china2011comparative), with most applications laying along this spectrum (@li2015plant, @kress2007two, @cbol2009dna). Particular challenges with the utilization of high-copy number sequences are associated with their rates of divergence, gene tree conflict, and hybridization (@coissac2016barcodes, @fazekas2009plant) Herein we have resolved major components of the problems of identifying plant material without diagnostic morphological character states using the Angiosperms353 (A353) Hyb-Seq probes (@johnson2019universal), and custom species sequence databases derived via species distribution modelling, and temporal filtering.

@@

  Currently the largest plant systematic endeavor ever undertaken, the Kew Plant and Fungal Tree of Life (PAFTOL), is approaching completion (@baker2021PAFTOL). This dataset will contain Hyb-Seq data from at least one species representing each genus in the plant kingdom using the popular A353 probes (@baker2021exploring), resulting in over 14,000 represented species. These publicly available data serve to provide a taxonomically comprehensive backbone for plant metabarcoding. Data from the 10kP project, which seeks to develop reference genomes from a phylogenetically diverse suite of plants will contribute many more records upon it's intended completion, now slated to be by 2030, similar projects which seek to sequence high amounts of genomes in regions e.g. the 'Darwin Tree of Life' are being undertaken which will contribute data for applicable to enormous spatial domains (@cheng2018tenkp, @darwin2022project, @lewin2022biogenome).

@@

## 3.2 | Microscopic Pollen identification
```{r, echo = F}
a <- c('Trifolium', 'Lupinus', 'Glyccrrhiza', 'Mitella', 'Geum')
b <- c('Lupinus', 'Lathyrus', 'Potentilla', 'Androsace', 'Bistorta', 'Vicia')
c <- c('Jeffersonia', 'Micranthes', 'Prunus', 'Delphinium', 'Androsace', 
       'Penstemon', 'Orthocarpus', 'Scuttelaria', 'Aquilegia', 'Castilleja', 'Draba')
d <- c('Salix', 'Boechera')

# Do we detect mertensia, asteraceae 1, asteraceae 2, in all molecular samples which they are present in the pollen samples of?

@@

Only 58 of these 388 species (n = 34.56897, Mdn = 31) were able to be compared to ground truth data from long term (1970 ? - 2016 ?) datasets.

@@

bee_obs <- read.csv(
  file.path('../data', files[grep('queen_observations', files)]) ) %>% 
  filter(caste == 'q') %>% 
  mutate(species = gsub('\\..*$', '', species)) %>% 
  select(site, doy, week, tot.obs.length, species, plant.species,
         resource.coll:pollen.collected, fl.switch.1:fl.switch.2.resource.coll) %>% 
  mutate(across(resource.coll:fl.switch.2.resource.coll, ~ na_if(.x, ''))) %>% 
  pivot_longer(c(plant.species, fl.switch.1, fl.switch.2), names_to = 'observation', values_to = 'plant.species') %>% 
  drop_na(plant.species) %>% 
  group_by(site, week, species, plant.species) %>% 
  count(name = 'interactions') %>% 
  left_join(., obs_time, by = c("site", "week")) %>% 
  mutate(period = case_when(
    week %in% 3:5 ~ 'Early',
    week %in% 6:8 ~ 'Mid',
    week %in% 9:11 ~ 'Late'
  ))

@@

```{r Most used resources by species, eval = F, echo =F}
